Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics

### 2020-01-15 paper/paper.Rmd
suggestions by Earl

# Remove contributors who did not provide any choices
bad_contribs <- d %>% group_by(contributor) %>% 
  summarise(sum0 = sum(choice)) %>% 
  filter(sum0 == 0) %>% 
  pull(contributor)

@@

In order to assess the effectiveness of the hexagon tile map, the lineup protocol [@GIIV,@BCHLLSW09] from visual inference procedures is employed. The procedures for doing a power comparison of competing plot designed, outlined in @GTPCCD, are followed. The approach mirrors classical statistical inference.

@@

Mean, standard deviations for lineups, by participant, ...

@@

Plots will be made in R (R Core Team 2019), with the ggplot2 package (Wickham 2016).

@@

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model. 
The R [@RCore] `glmer()` function in the [@lme4] package implements generalised linear mixed effect models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

@@

Computation will be done using R [@RCore], with the `lme4` package [@lme4].

@@

Responses from 95 participants were collected. Three participants provided no answers for any task, and their data was removed. Set A was evaluated by 42 participants, and 53 evaluated set B. This resulted in 1104 evaluations, corresponding to 92 subjects, each evaluating 12 lineups, that are analysed on accuracy and speed. The certainty and reasons of subjects in their answers is also examined.

@@

```{r detect-compare, fig.cap = "The detection rates achieved by participants are contrasted when viewing the four replicates of the three trend models. Each point shows the probability of detection for the lineup display, the facets separate the trend models hidden in the lineup. The points for the same data set shown in a choroleth or haxgon tile map display are linked to show the difference in the detection rate.", fig.height=4}
# Detectability rate for each lineup (image)
d_smry <- d %>% group_by(trend, type, replicate) %>%
  # pdetect measures the aggregated accuracy of the choices
  summarise(pdetect = length(detect[detect == 1])/length(detect)) %>%
  ungroup()

@@

# Plot summary
ggplot(d_smry, aes(x=type, y=pdetect, color = trend)) +
  geom_point(size = 2) +  
  geom_line(size = 1, aes(group = replicate)) +  
  facet_wrap(~trend) +
  scale_color_manual(values = trend_colors) +
  xlab("Type of areas visualized") +
  ylab("Detection rate") + 
  ylim(0,1) +
  guides(color = FALSE)
```

Table. \ref{tab:desc-stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard errors, the smallest standard deviation for all sets of replicates is the Three Cities trend model shown in a Choropleth display. This group of displays had a very small detection rate of 0.04. The mean detection rate for the Three Cities trend model shown as choropleth map lineups was also the smallest at 0.39.
The North-West to South-East (NW-SE) trend model unexpectedly had a higher mean detection rate for the hexagon tile map displays, but the difference in the means of detection rate was only 0.10.

@@

Table. \ref{tab:detect-glmer1} presents a summary of the generalised linear mixed effects model, testing the effect of plot type and trend model on the detection rate. The results support the summary from Fig. \ref{fig:detect-compare} and all terms are statistically significant despite the large standard deviations observed in Table. \ref{tab:desc-stats}. Overall, the hexagon tile map performs marginally better than the choropleth for all trend models, which is a pleasant surprise. Allowing for the interaction effect, the difference in detection rate decreases for population related displays for a choropleth map lineup, but increases for a hexagon tile map display.
The log odds of detection show in Table. \ref{tab:detect-glmer1} can be back transformed after taking the sum of all terms for the trend and type of display that are of interest.
For the NW-SE distribution, the predicted detection rate for the hexagon tile map display increases the predicted probability of detection to `r round(exp(0.461+0.0220)/(1+exp(0.461+0.0220)), 2)` from `r round(exp(0.0220)/(1+exp(0.0220)),2)` for choropleths, this is almost exactly the difference seen in the table of means and is significant only at the 0.05 level.

When a choropleth map display is used, the predicted detection rate for the Three Cities trend, `r round(exp(0.0220-3.43)/(1+exp(0.0220-3.43)),2)`; this is extremely low, especially compared to the NW-SE trend of `r round(exp(0.0220)/(1+exp(0.0220)),2)`.
When the All Cities trend is presented in a choropleth display the predicted probability of detection is `r round(exp(0.0220-1.35)/(1+exp(0.0220-1.35)),2)`.
The hexagon tile map has a substantially high detection rate for the display of a Three Cities trend `r round(exp(0.461+0.0220-3.43+2.46)/(1+exp(0.461+0.0220-3.43+2.46)),2)` and All Cities trend `r round(exp(0.461+0.0220-1.35+1.17)/(1+exp(0.461+0.0220-1.35+1.17)),2)`.

@@

ggplot(d %>% rename(Detected = detect_f), 
   aes(x = certainty, fill = Detected)) +  
  scale_fill_manual(values = detect_colors) +
  geom_bar() + facet_grid(type ~ trend) +
  theme(legend.position = "bottom") + xlab("Level of certainty")
```

@@

The reasons used when viewing choropleth displays varied more than the hexagon tile map reasons.
The hexagon tile map displays resulted in "Clusters of color" as the most common choice made by participants.

@@

The increasing population densities of capital cities despite large land area exacerbates the difference in the smallest and largest communities.
The population density structure of Australia can be considered similar to that of Canada, New Zealand and many others. Therefore, this display is not only relevant to Australia, but all nations or population distributions that experience densely populated cities separated by vast rural expanses.

>>>>

# Remove contributors who did not provide any choices
# Or an insufficient amount of responses
bad_contribs <- d %>% group_by(contributor) %>% 
  summarise(sum0 = sum(choice)) %>% 
  filter(sum0 < 13) %>% 
  pull(contributor)

@@

In order to assess the effectiveness of the hexagon tile map, the lineup protocol [@GIIV],[@BCHLLSW09] from visual inference procedures is employed. The procedures for doing a power comparison of competing plot designed, outlined in @GTPCCD, are followed. The approach mirrors classical statistical inference.

@@

Basic descriptive statistics were used to contrast the detection rate for the two types of displays. This was extended to consider the trend models, contrasting the mean and standard detection rate for each group, who had seen the different map display type for each trend model.

@@

Plots will be made in R [@R], with the \pkg{ggplot2} package [@ggplot2].

@@

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model. 
The R [@RCore] `glmer()` function in the \pkg{lme4} [@lme4] package implements generalised linear mixed effect models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

@@

Computation was performed using R [@RCore], with the `lme4` package [@lme4].

@@

Responses from 95 participants were collected. Five participants did not provide more than three answers for the twelve lineups, and their data was removed. Set A was evaluated by 42 participants, and 53 evaluated set B. This resulted in 1104 evaluations, corresponding to 92 subjects, each evaluating 12 lineups, that are analysed on accuracy and speed. The certainty and reasons of subjects in their answers is also examined.

@@

```{r detect-compare, fig.cap = "The detection rates achieved by participants are contrasted when viewing the four replicates of the three trend models. Each point shows the probability of detection for the lineup display, the facets separate the trend models hidden in the lineup. The points for the same data set shown in a choroleth or hexagon tile map display are linked to show the difference in the detection rate.", fig.height=4}
# Detectability rate for each lineup (image)
d_smry <- d %>% group_by(trend, type, replicate) %>%
  # pdetect measures the aggregated accuracy of the choices
  summarise(pdetect = length(detect[detect == 1])/length(detect)) %>%
  ungroup()

@@

# Plot summary
ggplot(d_smry, aes(x = type, y = pdetect, color = trend)) +
  geom_point(size = 2) +  
  geom_line(size = 1, aes(group = replicate)) +  
  facet_wrap(~trend) +
  scale_color_manual(values = trend_colors) +
  xlab("Type of areas visualized") +
  ylab("Detection rate") + 
  ylim(0,1) +
  guides(color = FALSE)
```

Table. \ref{tab:desc-stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard errors, the smallest standard deviation for all sets of replicates is the Three Cities trend model shown in a Choropleth display. This group of displays had a very small detection rate of 0.04. The mean detection rate for the Three Cities trend model shown as choropleth map lineups was also the smallest at 0.40.
The North-West to South-East (NW-SE) trend model unexpectedly had a higher mean detection rate for the hexagon tile map displays, but the difference in the means of detection rate was only 0.10.

@@

Table. \ref{tab:detect-glmer1} presents a summary of the generalised linear mixed effects model, testing the effect of plot type and trend model on the detection rate. The results support the summary from Fig. \ref{fig:detect-compare} and all parameters are statistically significant despite the large standard deviations observed in Table. \ref{tab:desc-stats}. Overall, the hexagon tile map performs marginally better than the choropleth for all trend models, which is a pleasant surprise. Allowing for the interaction effect, the difference in detection rate decreases for population related displays for a choropleth map lineup, but increases for a hexagon tile map display.
The log odds of detection show in Table. \ref{tab:detect-glmer1} can be back transformed after taking the sum of all terms for the trend and type of display that are of interest.
For the NW-SE distribution, the predicted detection rate for the hexagon tile map display increases the predicted probability of detection to `r round(exp(0.46+0.07)/(1+exp(0.46+0.07)), 2)` from `r round(exp(0.07)/(1+exp(0.07)),2)` for choropleths, this is almost exactly the difference seen in the table of means and is significant only at the 0.05 level.

When a choropleth map display is used, the predicted detection rate for the Three Cities trend, `r round(exp(0.07-3.41)/(1+exp(0.07-3.41)),2)`; this is extremely low, especially compared to the NW-SE trend of `r round(exp(0.07)/(1+exp(0.07)),2)`.
When the All Cities trend is presented in a choropleth display the predicted probability of detection is `r round(exp(0.07-1.34)/(1+exp(0.07-1.34)),2)`.
The hexagon tile map has a substantially high detection rate for the display of a Three Cities trend `r round(exp(0.46+0.07-3.41+2.44)/(1+exp(0.46+0.07-3.41+2.44)),2)` and All Cities trend `r round(exp(0.46+0.07-1.34+1.16)/(1+exp(0.46+0.07-1.34+1.16)),2)`.

@@

d %>% 
  mutate(Detected = factor(detect_f, 
    levels = c("Detected? Yes", "Detected? No"), 
    labels = c("Yes", "No"))) %>% 
ggplot(aes(x = certainty, fill = Detected)) +  
  scale_fill_manual(values = detect_f_colors) +
  geom_bar() + facet_grid(type ~ trend) +
  theme(legend.position = "bottom") + xlab("Level of certainty")
```

@@

The reasons chosen by participants from the list provided to them varied more when viewing choropleth displays than the hexagon tile map.
The hexagon tile map displays resulted in "Clusters of color" as the most common choice made by participants.

@@

The increasing population densities of capital cities despite large land area exacerbates the difference in the smallest and largest communities.
The population density structure of Australia can be considered similar to that of Canada, New Zealand and many other countries. Therefore, this display is not only relevant to Australia, but all nations or population distributions that experience densely populated cities separated by vast rural expanses.

### 2020-01-15 paper/paper.tex
suggestions by Earl

In order to assess the effectiveness of the hexagon tile map, the lineup protocol {[}\protect\hyperlink{ref-GIIV}{3}{]} from visual inference procedures is employed. The procedures for doing a power comparison of competing plot designed, outlined in {[}\protect\hyperlink{ref-GTPCCD}{13}{]}, are followed. The approach mirrors classical statistical inference.

@@

Mean, standard deviations for lineups, by participant, \ldots{}

@@

Plots will be made in R (R Core Team 2019), with the ggplot2 package (Wickham 2016).

@@

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model.
The R {[}\protect\hyperlink{ref-RCore}{18}{]} \texttt{glmer()} function in the {[}\protect\hyperlink{ref-lme4}{19}{]} package implements generalised linear mixed effect models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

@@

Computation will be done using R {[}\protect\hyperlink{ref-RCore}{18}{]}, with the \texttt{lme4} package {[}\protect\hyperlink{ref-lme4}{19}{]}.

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/detect-compare-1.pdf}
\caption{\label{fig:detect-compare}The detection rates achieved by participants are contrasted when viewing the four replicates of the three trend models. Each point shows the probability of detection for the lineup display, the facets separate the trend models hidden in the lineup. The points for the same data set shown in a choroleth or haxgon tile map display are linked to show the difference in the detection rate.}
\end{figure}

Table. \ref{tab:desc-stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard errors, the smallest standard deviation for all sets of replicates is the Three Cities trend model shown in a Choropleth display. This group of displays had a very small detection rate of 0.04. The mean detection rate for the Three Cities trend model shown as choropleth map lineups was also the smallest at 0.39.
The North-West to South-East (NW-SE) trend model unexpectedly had a higher mean detection rate for the hexagon tile map displays, but the difference in the means of detection rate was only 0.10.

@@

\caption{\label{tab:desc-stats}The rate of detection for each trend model has been calculate for the choropleth and hexagon tile map displays. The associated standard errors are also included.}
\centering
\begin{tabular}[t]{lccc}
\toprule
Type & NW-SE & Three Cities & All Cities\\
\midrule
Choro. & 0.51 & 0.04 & 0.23\\
 & (0.50) & (0.19) & (0.42)\\
\addlinespace
Hex. & 0.61 & 0.39 & 0.57\\
 & (0.49) & (0.49) & (0.50)\\
\bottomrule
\end{tabular}
\end{table}

Table. \ref{tab:detect-glmer1} presents a summary of the generalised linear mixed effects model, testing the effect of plot type and trend model on the detection rate. The results support the summary from Fig. \ref{fig:detect-compare} and all terms are statistically significant despite the large standard deviations observed in Table. \ref{tab:desc-stats}. Overall, the hexagon tile map performs marginally better than the choropleth for all trend models, which is a pleasant surprise. Allowing for the interaction effect, the difference in detection rate decreases for population related displays for a choropleth map lineup, but increases for a hexagon tile map display.
The log odds of detection show in Table. \ref{tab:detect-glmer1} can be back transformed after taking the sum of all terms for the trend and type of display that are of interest.
For the NW-SE distribution, the predicted detection rate for the hexagon tile map display increases the predicted probability of detection to 0.62 from 0.51 for choropleths, this is almost exactly the difference seen in the table of means and is significant only at the 0.05 level.

When a choropleth map display is used, the predicted detection rate for the Three Cities trend, 0.03; this is extremely low, especially compared to the NW-SE trend of 0.51.
When the All Cities trend is presented in a choropleth display the predicted probability of detection is 0.21.
The hexagon tile map has a substantially high detection rate for the display of a Three Cities trend 0.38 and All Cities trend 0.58.

@@

\caption{\label{tab:detect-glmer1}The model output for the generalised linear mixed effect model for detection rate. This model considers the type of display, the trend model hidden in the data plot, and accounts for contributor performance.}
\centering
\begin{tabular}[t]{rrlrr}
\toprule
Term & Est. & Sig. & Std. Error & P val\\
\midrule
Intercept & 0.02 & $^{ }$ & 0.17 & 0.90\\
Hex. & 0.46 & $^{*}$ & 0.22 & 0.04\\
\addlinespace
Three Cities & -3.43 & $^{***}$ & 0.42 & 0.00\\
All Cities & -1.35 & $^{***}$ & 0.24 & 0.00\\
\addlinespace
Hex:Three Cities & 2.46 & $^{***}$ & 0.47 & 0.00\\
Hex:All Cities & 1.17 & $^{***}$ & 0.33 & 0.00\\
\bottomrule
\end{tabular}
\end{table}

@@

The reasons used when viewing choropleth displays varied more than the hexagon tile map reasons.
The hexagon tile map displays resulted in ``Clusters of color'' as the most common choice made by participants.

>>>>

In order to assess the effectiveness of the hexagon tile map, the lineup protocol {[}\protect\hyperlink{ref-GIIV}{3}{]},{[}\protect\hyperlink{ref-BCHLLSW09}{12}{]} from visual inference procedures is employed. The procedures for doing a power comparison of competing plot designed, outlined in {[}\protect\hyperlink{ref-GTPCCD}{13}{]}, are followed. The approach mirrors classical statistical inference.

@@

Basic descriptive statistics were used to contrast the detection rate for the two types of displays. This was extended to consider the trend models, contrasting the mean and standard detection rate for each group, who had seen the different map display type for each trend model.

@@

Plots will be made in R {[}{\textbf{???}}{]}, with the \pkg{ggplot2} package {[}\protect\hyperlink{ref-ggplot2}{14}{]}.

@@

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model.
The R {[}\protect\hyperlink{ref-RCore}{18}{]} \texttt{glmer()} function in the \pkg{lme4} {[}\protect\hyperlink{ref-lme4}{19}{]} package implements generalised linear mixed effect models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

@@

Computation was performed using R {[}\protect\hyperlink{ref-RCore}{18}{]}, with the \texttt{lme4} package {[}\protect\hyperlink{ref-lme4}{19}{]}.

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/detect-compare-1.pdf}
\caption{\label{fig:detect-compare}The detection rates achieved by participants are contrasted when viewing the four replicates of the three trend models. Each point shows the probability of detection for the lineup display, the facets separate the trend models hidden in the lineup. The points for the same data set shown in a choroleth or hexagon tile map display are linked to show the difference in the detection rate.}
\end{figure}

Table. \ref{tab:desc-stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard errors, the smallest standard deviation for all sets of replicates is the Three Cities trend model shown in a Choropleth display. This group of displays had a very small detection rate of 0.04. The mean detection rate for the Three Cities trend model shown as choropleth map lineups was also the smallest at 0.40.
The North-West to South-East (NW-SE) trend model unexpectedly had a higher mean detection rate for the hexagon tile map displays, but the difference in the means of detection rate was only 0.10.

@@

\caption{\label{tab:desc-stats}The rate of detection for each trend model has been calculate for the choropleth and hexagon tile map displays. The associated standard errors are also included.}
\centering
\begin{tabular}[t]{lccc}
\toprule
Type & NW-SE & Three Cities & All Cities\\
\midrule
Choro. & 0.52 & 0.04 & 0.23\\
 & (0.50) & (0.19) & (0.42)\\
\addlinespace
Hex. & 0.62 & 0.4 & 0.58\\
 & (0.49) & (0.49) & (0.49)\\
\bottomrule
\end{tabular}
\end{table}

Table. \ref{tab:detect-glmer1} presents a summary of the generalised linear mixed effects model, testing the effect of plot type and trend model on the detection rate. The results support the summary from Fig. \ref{fig:detect-compare} and all parameters are statistically significant despite the large standard deviations observed in Table. \ref{tab:desc-stats}. Overall, the hexagon tile map performs marginally better than the choropleth for all trend models, which is a pleasant surprise. Allowing for the interaction effect, the difference in detection rate decreases for population related displays for a choropleth map lineup, but increases for a hexagon tile map display.
The log odds of detection show in Table. \ref{tab:detect-glmer1} can be back transformed after taking the sum of all terms for the trend and type of display that are of interest.
For the NW-SE distribution, the predicted detection rate for the hexagon tile map display increases the predicted probability of detection to 0.63 from 0.52 for choropleths, this is almost exactly the difference seen in the table of means and is significant only at the 0.05 level.

When a choropleth map display is used, the predicted detection rate for the Three Cities trend, 0.03; this is extremely low, especially compared to the NW-SE trend of 0.52.
When the All Cities trend is presented in a choropleth display the predicted probability of detection is 0.22.
The hexagon tile map has a substantially high detection rate for the display of a Three Cities trend 0.39 and All Cities trend 0.59.

@@

\caption{\label{tab:detect-glmer1}The model output for the generalised linear mixed effect model for detection rate. This model considers the type of display, the trend model hidden in the data plot, and accounts for contributor performance.}
\centering
\begin{tabular}[t]{rrlrr}
\toprule
Term & Est. & Sig. & Std. Error & P val\\
\midrule
Intercept & 0.07 & $^{ }$ & 0.16 & 0.67\\
Hex. & 0.46 & $^{*}$ & 0.22 & 0.04\\
\addlinespace
Three Cities & -3.41 & $^{***}$ & 0.42 & 0.00\\
All Cities & -1.34 & $^{***}$ & 0.24 & 0.00\\
\addlinespace
Hex:Three Cities & 2.44 & $^{***}$ & 0.47 & 0.00\\
Hex:All Cities & 1.16 & $^{***}$ & 0.33 & 0.00\\
\bottomrule
\end{tabular}
\end{table}

@@

The reasons chosen by participants from the list provided to them varied more when viewing choropleth displays than the hexagon tile map.
The hexagon tile map displays resulted in ``Clusters of color'' as the most common choice made by participants.
